# Regression and model validation

Load the wrangled data and take a look at it.
```{r}
learning2014 <- read.table("data/learning2014.tsv", sep = "\t")
str(learning2014)
```

Plot the data.
```{r}
library(ggplot2)

# initialize plot with data and aesthetic mapping
p1 <- ggplot(learning2014, aes(x = attitude, y = points, col = gender))

# define the visualization type (points)
p2 <- p1 + geom_point()

# add a regression line
p3 <- p2 + geom_smooth(method = "lm")

# add a main title and draw the plot
p4 <- p3 + ggtitle("Student's attitude versus exam points")

# draw the plot
p4
```

There is a positive correlation between attitude and points.
There is no obvious difference between the two genders.
Interestingly, there are only two genders.

Now, plot all possible pairs of variables as scatter plots:

```{r}
pairs(learning2014[-1])
```

This is not very informative, since the plots are very small and there is no regression line to help us imagine which way the correlation goes.

Drawing more advanced plots:

```{r}
library(GGally)
library(ggplot2)

# create a more advanced plot matrix with ggpairs()
p <- ggpairs(learning2014, mapping = aes(col = gender, alpha = .3), lower = list(combo = wrap("facethist", bins = 20)))

# draw the plot
p
```

Age seems to follow a Poisson distribution, probably because the subjects are students.
The rest of the variables seem to have a normal distribution, as expected.

Apparently, attitude and points have the highest correlation by far.

In addition, there is a negative correlation between surf and deep.

There is a slight negative correlation between surf and points, surf and age, and surf and stra, as well as a positive correlation between stra and points.

Now, a regression analysis using the three variables that had the highest individual correlation with points:
```{r}
# create a regression model with three explanatory variables
my_model2 <- lm(points ~ attitude + stra + surf, data = learning2014)

# print out a summary of the model
summary(my_model2)
```

There seems to be a statistically significant relationship between the chosen variables.
Here are the diagnostic plots:
```{r}
par(mfrow = c(2,2))
plot(my_model2, which = c(1, 2, 5))
```

Nothing looks out of the ordinary in these diagnostic plots.
There is no reason not to trust the hypothesis that all three chosen variables -- attitude, stra, and surf, are explanatory variables for points.
The correlation of course is not incredibly strong.
