# Analysis of longitudinal data

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
```

## BPRS dataset
In the BPRS dataset 40 male subjects were randomly assigned to one of two treatment groups.
Each subject was rated on the brief psychiatric rating scale (BPRS) measured before treatment began (week 0) and then at weekly intervals for eight weeks.
The BPRS assesses the level of 18 symptom constructs such as hostility, suspiciousness, hallucinations and grandiosity;
each of these is rated from 1 (not present) to 7 (extremely severe).
The scale is used to evaluate patients suspected of having schizophrenia.

Read the dataset:
```{r}
BPRSL <- read.csv("data/BPRSL.csv")
BPRSL$treatment <- factor(BPRSL$treatment)
BPRSL$subject <- factor(BPRSL$subject)
str(BPRSL)
```

Plot the `bprs` values over time for each individual by treatment group: 
```{r}
ggplot(BPRSL, aes(x = week, y = bprs, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") +
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))
```

The BPRS score and the variability of the score in both treatment groups decrease over time.

Standardize the scores for each time point:
subtract the average `bprs` for all values and divide by the standard deviation:
```{r}
BPRSL <- BPRSL %>%
  group_by(week) %>%
  mutate(stdbprs = scale(bprs)) %>%
  ungroup()

glimpse(BPRSL)
```

Plot the standardized values:
```{r}
ggplot(BPRSL, aes(x = week, y = stdbprs, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  scale_y_continuous(name = "standardized bprs")
```

Plot the average `bprs` for each time point for the two different treatment groups, and add the standard error of the means to the plots:

$$se = \frac{sd(x)}{\sqrt{n}}$$

```{r}
n <- BPRSL$week %>% unique() %>% length()
BPRSS <- BPRSL %>%
  group_by(treatment, week) %>%
  summarise(mean = mean(bprs), se = sd(bprs)/sqrt(n) ) %>%
  ungroup()
glimpse(BPRSS)
ggplot(BPRSS, aes(x = week, y = mean, linetype = treatment, shape = treatment)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2)) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.8,0.8)) +
  scale_y_continuous(name = "average(bprs) +/- sem(bprs)")
```

The averaged profiles overlap completely when taking into account the standard errors of the mean.
This suggests there is only a small difference between the treatment groups.

Compare the average `bprs` values between the treatment groups on weeks 1 to 8 by plotting the distribution of the averaged `bprs` values for the two groups:
```{r}
BPRSL8S <- BPRSL %>%
  filter(week > 0) %>%
  group_by(treatment, subject) %>%
  summarise(mean=mean(bprs)) %>%
  ungroup()
glimpse(BPRSL8S)
ggplot(BPRSL8S, aes(x = treatment, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(bprs), weeks 1-8")
```

There is an outlier in group 2 with a `bprs` value of over 70.
Remove it so that it does not bias the results:
```{r}
BPRSL8S1 <- BPRSL %>%
  filter(week > 0) %>%
  group_by(treatment, subject) %>%
  summarise(mean=mean(bprs)) %>%
  ungroup() %>%
  filter(mean < 70)
ggplot(BPRSL8S1, aes(x = treatment, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(bprs), weeks 1-8")
```

Looking at the plot one might think that the average `brps` is lower for treatment group 2, but the variation of the mean inside the groups is larger.

Perform a _t_-test comparing the average `bprs` values between the treatment groups:
```{r}
t.test(mean ~ treatment, data = BPRSL8S1, var.equal = TRUE)
```

There is no statistically significant difference between the groups.

The baseline `bprs` value might be correlated with the chosen summary measure.
Add that to the model to see if that will affect the difference between the treatment groups:
```{r}
baseline <- BPRSL %>% 
  filter(week == 0) %>% 
  rename(baseline=bprs) %>%
  dplyr::select(one_of(c("treatment", "subject", "baseline")))
BPRSL8S2 <- BPRSL8S %>%
  left_join(baseline)
fit <- lm(mean ~ treatment + baseline, data = BPRSL8S2)
anova(fit)
```

The baseline `bprs` values is strongly associated with the `bprs` values taken after treatment has begun;
still, there is no evidence of a treatment difference even after conditioning on the baseline value.

## RATS dataset
The RATS dataset comes from a nutrition study conducted in three groups of... you guessed it, **rats**.
The groups were on different diets.
The body weight of each animal was recorder weekly over a period of 9 weeks (except in week 7: it was recorder twice).

Read in the dataset:
```{r}
RATSL <- read.csv("data/RATSL.csv")
RATSL$ID <- factor(RATSL$ID)
RATSL$Group <- factor(RATSL$Group)
str(RATSL)
```

Plot the RATSL dataset:
```{r}
ggplot(RATSL, aes(x = Time, y = Weight, group = ID)) +
  geom_line(aes(linetype = Group)) +
  scale_x_continuous(name = "Time (days)", breaks = seq(0, 60, 10)) +
  scale_y_continuous(name = "Weight (grams)") +
  theme(legend.position = "top")
```

The weight of the rats in group 1 is lower at the start of the follow-up compared to the rats in groups 2 and 3, and stays lower during the follow-up.

### Linear regression model
Fit a linear regression model where `Weight` is the outcome and `Group` and `Time` are the explanatory variables:

$$Weight \sim Group + Time$$.

We are making the (highly unlikely!) assumption that the consecutive weights of the same animal are independent:
```{r}
RATS_reg <- lm(Weight ~ Time + Group, data=RATSL)
summary(RATS_reg)
```

Weight is statistically significantly higher in groups 2 and 3 compared to group 1.
The regression coefficient of time is $<1$ and statistically significant:
the weight of the animals goes down during the follow-up.

### Linear mixed-effects models
Fit a random intercept model using the same two explanatory variables, `Time` and `Group`.
To allow the rats to have a different weight at the start of the follow-up we use the identity of each rat as the random effect:
```{r message=FALSE}
library(lme4)
RATS_ref <- lmer(Weight ~ Time + Group + (1 | ID), data = RATSL, REML = FALSE)
summary(RATS_ref)
```

Even after allowing different weight of the animals at the start of the follow-up, the animals in groups 2 and 3 are heavier than those group 1,
and the weight of the animals decreases during the follow-up.

Add random slope to the model of the rat growth data.
Using a random intercept and random slope model allows the linear regression fits for each animal to differ in intercept and slope.
This way we take into account that the rats start with different weights, and their weights might change over time at different rates,
as well as analyse the effect of time in general:
```{r}
RATS_ref1 <- lmer(Weight ~ Time + Group + (Time | ID), data = RATSL, REML = FALSE)
summary(RATS_ref1)
```

The animals in groups 2 and 3 are heavier than those in group 1;
the weight decreases on average over time.

Compare the random intercept and random intercept and slope models by performing a likelihood ratio test:
```{r}
anova(RATS_ref1, RATS_ref)
```

The _p_-value is highly significant and the log-likelihood of the random intercept and random slope is greater than that of the random slope model.
This suggests that it fits the data better: the fit is better the closer to 0 the log-likelihood of the model is.

To test if the growth profiles of the rats differ between the groups, fit a random intercept and slope model that allows for a `Group` × `Time` interaction:
```{r}
RATS_ref2 <- lmer(Weight ~ Time + Group + Time * Group + (Time | ID), data = RATSL, REML = FALSE)
summary(RATS_ref2)
```

The interaction of time and weight is stronger in groups 2 and 3 than in group 1:
the animals gain weight faster.

Compare the random intercept and random slope model to the random intercept and random slope with `Time` and `Weight` interaction using ANOVA: 
```{r}
anova(RATS_ref2, RATS_ref1)
```

The model with `Time` and `Weight` interaction fits the data better, based on the smaller log-likelihood of the model.
The difference is statistically significant.

Visualize the observed weights and the fitted values for weight from the last model:
```{r}
ggplot(RATSL, aes(x = Time, y = Weight, group = ID)) +
  geom_line(aes(linetype = Group)) +
  scale_x_continuous(name = "Time (days)", breaks = seq(0, 60, 20)) +
  scale_y_continuous(name = "Observed weight (grams)") +
  theme(legend.position = "top")
Fitted <- fitted(RATS_ref2)
RATSL <- mutate(RATSL, Fitted=Fitted)
ggplot(RATSL, aes(x = Time, y = Fitted, group = ID)) +
  geom_line(aes(linetype = Group)) +
  scale_x_continuous(name = "Time (days)", breaks = seq(0, 60, 20)) +
  scale_y_continuous(name = "Fitted weight (grams)") +
  theme(legend.position = "top")
```
